### 2023-12-26 sections/3-datasets.tex
some changes to datasets and results

We benchmark on a variety of datasets from the ann-benchmarks suite~\cite{Aumller2018ANNBenchmarksAB}, along with the appropriate distance function.
Table~\ref{table:datasets:summary} summarizes the properties of these datasets.

We compared against FAISS~\cite{johnson2019billion} (IVF and HNSW) and a BLAS-accelerated implementation of na\"ive search as reported in~\cite{johnson2019billion}.

The benchmarks for Fashion-MNIST, Glove-25, Sift and Random (and for their synthetic augmentations) were conducted on an Amazon AWS (EC2) \texttt{r6i.16xlarge} instance, with a 64-core Intel Xeon Platinum 8375C CPU 2.90GHz processor, 512GB RAM.
This is the same configuration used in~\cite{Aumller2018ANNBenchmarksAB}.
The OS kernel was Ubuntu 22.04.3-Ubuntu SMP.
The Rust compiler was Rust 1.72.0, and the Python interpreter version was 3.9.16.

@@

In addition to benchmarks on datasets from the ANN Benchmarks suite, we also benchmarked on synthetic augmentations of these real datasets, using the process described in \ref{subsec:methods:synthetic-data}. 
In particular, we use a noise tolerance $\epsilon = 0.01$ and explore the scaling behavior as the cardinality multiplier (refered to as ``scale'' in Table~\ref{table:results:ann-fashion} {\color{red} TODO: merge the tables into one and confirm this reference is still correct after the merge}) increases.

For the Random dataset, we used a cardinality of 1,000,000 and a dimensionality of 128 to match the Sift dataset.
This benchmark allows us to isolate the effect of a manifold structure (which we expect to be absent in a purely random dataset) on the performance of the CAKES' algorithms.

\subsection{Silva}
\label{subsec:silva}

Silva 18S~\cite{10.1093/nar/gks1219} contains ribosomal DNA sequences of 2,224,640 genomes with an aligned length of 50,000 letters.
We use Levenshtein distance on unaligned sequences as the distance function for this dataset.
{\color{red} TODO: Define and/or cite Levenshtein distance.}
% I think this needs another sentence but I'm not sure what else to say

\subsection{Radio ML}
\label{subsec:radioml}

The RadioML dataset contains samples of both live and synthetically generated signal captures of different modulation modes over a range of SNR levels.
Specifically, it is comprised of 24 modulation modes at 26 different SNR levels ranging from -20 dB to 30 dB, with 4,096 samples at each modulation mode and SNR level~\cite{oshea2018radioml}.
Thus, it contains $24 \cdot 26 \cdot 4096 = 2,555,504$ samples in total.
Each sample is a 1,024-dimensional complex-valued vector, representing a signal capture (a time series of complex-valued numbers).
We used a subset of this dataset, containing only the 97,920 samples at 10dB SNR.
We use Dynamic Time Warping as the distance function for this dataset.
{\color{red} TODO: Define and/or cite Dynamic Time Warping distance.}

>>>>

We benchmark on a variety of datasets from the ANN-benchmarks suite~\cite{Aumller2018ANNBenchmarksAB}, along with the appropriate distance function.
Table~\ref{table:datasets:summary} summarizes the properties of these datasets.

The benchmarks for Fashion-Mnist, Glove-25, and Sift (and for their synthetic augmentations) were conducted on an Amazon AWS (EC2) \texttt{r6i.16xlarge} instance, with a 64-core Intel Xeon Platinum 8375C CPU 2.90GHz processor, 512GB RAM.
This is the same configuration used in~\cite{Aumller2018ANNBenchmarksAB}.
The OS kernel was Ubuntu 22.04.3-Ubuntu SMP.
The Rust compiler was Rust 1.72.0, and the Python interpreter version was 3.9.16.

@@

In addition to benchmarks on datasets from the ANN Benchmarks suite, we also benchmarked on synthetic augmentations of these real datasets, using the process described in \ref{subsec:methods:synthetic-data}. 
In particular, we use a noise tolerance $\epsilon = 0.01$ and explore the scaling behavior as the cardinality multiplier (referred to as ``scale'' in Table~\ref{table:results:ann-fashion} {\color{red} TODO: merge the tables into one and confirm this reference is still correct after the merge}) increases.

We also benchmarked on purely random (i.e., not an augmented version of a known dataset) datasets of various cardinalities. 
For this Random dataset, we used a base cardinality of 1,000,000 and a dimensionality of 128 to match the Sift dataset.
This benchmark allows us to isolate the effect of a manifold structure (which we expect to be absent in a purely random dataset) on the performance of the CAKES' algorithms.

\subsection{Silva}
\label{subsec:silva}
To showcase the use of CAKES with a more exotic distance function, we also benchmarked on the Silva dataset.
Silva 18S~\cite{10.1093/nar/gks1219} contains ribosomal DNA sequences of 2,224,640 genomes with an aligned length of 50,000 letters.
We use Levenshtein distance on unaligned sequences as the distance function for this dataset.
{\color{red} TODO: Define and/or cite Levenshtein distance.}
% I think this needs another sentence but I'm not sure what else to say

\subsection{Radio ML}
\label{subsec:radioml}
To provide another example of use of CAKES with a more exotic distance function, we benchmarked on the RadioML dataset.
The RadioML dataset contains samples of both live and synthetically generated signal captures of different modulation modes over a range of SNR levels.
Specifically, it is comprised of 24 modulation modes at 26 different SNR levels ranging from -20 dB to 30 dB, with 4,096 samples at each modulation mode and SNR level~\cite{oshea2018radioml}.
Thus, it contains $24 \cdot 26 \cdot 4096 = 2,555,504$ samples in total.
Each sample is a 1,024-dimensional complex-valued vector, representing a signal capture (a time series of complex-valued numbers).
We used a subset of this dataset, containing only the 97,920 samples at 10dB SNR.
We use Dynamic Time Warping as the distance function for this dataset.
{\color{red} TODO: Define and/or cite Dynamic Time Warping distance.}

\subsection{Other Algorithms}
We benchmarked CAKES's algorithms against the state-of-the-art similarity search algorithms HNSW and ANNOY, as well as on FAISS-IVF and two different implementations of linear search.
In particular, we use FAISS-Flat, a Python implementation of linear search, to provide ground-truth values for recall of HNSW, ANNOY, and FAISS-IVF.
We use our own Rust implementation of linear search to provide ground-truth values for recall of CAKES's algorithms.
We plot the throughput of these two implementations of linear search as a reference point for CAKES's algorithms' throughput in Section~\ref{subsec:scaling-behavior-results}.

### 2023-12-26 sections/4-results.tex
some changes to datasets and results

Figures~\ref{fig:results:fashion-mnist-scaling}, ~\ref{fig:results:glove-25-scaling}, and ~\ref{fig:results:sift-scaling} show the scaling behavior of CAKES algorithms and existing algorithms on augmented versions of the following ann-benchmark datasets: fashion-mnist under Euclidean distance, glove-25 under cosine distance, and sift under Euclidean distance.
Figure~\ref{fig:results:random-scaling} shows the scaling behavior of CAKES algorithms and existing algorithms on a completely randomly generated dataset with same cardinality and dimensionality as sift (1 million points in 128 dimensions).
Figures~\ref{fig:results:silva-scaling} and ~\ref{fig:results:radioml-scaling} show the scaling behavior of CAKES on Silva and RadioML respectively.
The horizontal axis in each figure shows the cardinality of the augmented dataset with synthetic points (see Section \ref{subsec:methods:synthetic-data}).
The left-most point on each line is on the original dataset without any synthetic augmentation.
The vertical axis denotes throughput in queries per second.
Both axes are on a logarithmic scale.
In this section, we report results only for $k$-NN search with $k = 10$, but similar plots for $k=100$ can be found in the Supplement.
For HNSW and Annoy, we report the recall for each measurement in the plots. For algorithms which exhibit recall greater than $0.9995$, we do not report the recall in the plots, but we do report it in the tables below.

Tables \ref{table:results:ann-fashion}, \ref{table:results:ann-glove-25}, \ref{table:results:ann-sift}, and \ref{table:results:ann-random} show the throughput and recall of CAKES's algorithms at each cardinality for  Fashion-mnist, Glove-25, Sift, and a random dataset respectively. Though the scaling plots above present results for each of CAKES's three algorithms separately, the results in the CAKES column in these tables represent the highest throughput CAKES algorithm at that dataset and cardinality only. 
Given that we redo the auto-tuning process described in Section~\ref{subsec:methods:auto-tuning} for each augmented dataset, these values are an accurate representation of CAKES's performance on each dataset and cardinality multiplier. 
When reporting recall, we use 1.000* to denote that the recall is imperfect, but rounds to 1.000 when we consider only three decimal places.

In Figure~\ref{fig:results:fashion-mnist-scaling}, which shows results for Fashion-mnist, we observe that as cardinality increases, the CAKES algorithms (GreedySieve in blue, Repeated $\rho$-NN in green, and Sieve in purple) exhibit higher throughput than na\"{i}ve linear search (plotted in orange).
In particular, we see that GreedySieve and Sieve begin to exhibit sublinear time performance at a cardinality near $10^5$ (i.e., at a multiplier of about {\color{red} add value} times the original dataset's cardinality). 
We also observe that CAKES's algorithms outperform FAISS-IVF (plotted in pink) for all cardinalities larger than the original dataset's cardinality.
Though state-of-the-art algorithms HNSW (plotted in gray), and Annoy (plotted in yellow) have higher throughput than CAKES for all cardinalities, we note that CAKES always exhibits perfect recall while HNSW and Annoy exhibit much lower recall, as shown in Table~\ref{table:results:ann-fashion}.
For example, at a multiplier as low as eight, HNSW and Annoy have recall of 0.525 and 0.857 respectively,
and this recall gets worse as the multiplier increases.
Meanwhile, CAKES exhibits perfect recall.
Amongst the CAKES algorithms, GreedySieve consistently performs best on this dataset, exhibiting sublinear time performance with increasing multipliers.

With glove-25 (Figure~\ref{fig:results:glove-25-scaling}), we observe that each of CAKES' algorithms outperforms linear search at some cardinality.
Specifically, all three of CAKES' algorithms outperform linear search at a cardinality just slightly more than the original dataset's cardinality, $10^6$.
Similar to the results for Fashion-mnist, we see that CAKES's algorithms outperform FAISS-IVF for all cardinalities greater than $10^7$.
We see once again that HNSW and Annoy have higher throughput than CAKES's algorithms for all cardinality multipliers, CAKES's algorithms exhibit better recall at all multipliers, as shown in Table~\ref{table:results:ann-glove-25}.
In particular, at a multiplier as low as two, HNSW and Annoy have recall as low as 0.607 and 0.832 respectively, and this recall continues to decrease as we increase the multiplier.
In contrast, CAKES exhibits perfect recall at all multipliers.
On this dataset, unlike with fashion-mnist, Repeated $\rho$-NN is consistently the best-performing CAKES algorithm.

In Figure~\ref{fig:results:sift-scaling}, which displays the scaling behavior of algorithms on the Sift dataset, we observe that while linear search performs best for lower cardinalities, our algorithms begin to outperform our implementation of linear search and FAISS-IVF at a cardinality near $10^7$.
For this dataset, Sieve, is the highest-throughput CAKES algorithm for all cardinalities.
Similar to the results with Fashion-mnist and Glove-25, we observe that while HNSW and Annoy exhibit higher throughput than CAKES's algorithms for all cardinality multipliers, CAKES has better recall.
Table~\ref{table:results:ann-sift} shows that even with a multiplier of 1 (i.e., the original dataset), HNSW and Annoy have recall as low as 0.782 and 0.686 respectively, while CAKES exhibits perfect recall.

With the random dataset, as seen in Figure~\ref{fig:results:random-scaling}, we observe that the three CAKES algorithms exhibit the worst throughput out of all algorithms, at all cardinalities. 
As with the real datasets, HNSW and Annoy exhibit the highest throughput of any of the algorithms, and CAKES exhibits perfect recall at all cardinalities. 
HNSW and Annoy exhibit \emph{much} lower recall on this random dataset than on the real datasets; 
in particular, with a multiplier of 1, HNSW and Annoy have recall as low as 0.060 and 0.028 respectively, as reported in Table~\ref{table:results:ann-random}.

For Silva and RadioML, we benchmarked only CAKES's algorithms.
Due to the massive sizes of these datasets, we took random sub-samples of the dataset with lower cardinalities for our benchmarks, rather than augmented versions of the dataset.
With Silva, as shown in Figure~\ref{fig:results:silva-scaling}, we observe that for all algorithms, throughput seems to linearly decrease as cardinality increases, but that it seems to begin levelling off near cardinality $10^5$. Until cardinality near $10^4$, GreedySieve is the highest-throughput CAKES algorithm, but for cardinalities greater than that, Repeated $\rho$-NN is the highest-throughput CAKES algorithm.

For RadioML, as shown in Figure~\ref{fig:results:radioml-scaling}, we observe that throughput declines nearly linearly the CAKES algorithms exhibit nearly indistinguishable performance for all cardinalities.

\begin{figure}
\begin{subfigure}[b]{0.47\textwidth}
\includegraphics[width=0.95\textwidth]{plots/fashion-mnist-knn-10.png}
\subcaption{Fashion-mnist for $k=10$.}
\label{fig:results:fashion-mnist-scaling}
\end{subfigure}%
\begin{subfigure}[b]{0.47\textwidth}
\includegraphics[width=0.95\textwidth]{plots/glove-25-knn-10.png}
\subcaption{Glove-25 for $k=10$.  }
\label{fig:results:glove-25-scaling}
\end{subfigure}%
\vspace{1em}
\\
\begin{subfigure}[b]{0.47\textwidth}
\includegraphics[width=0.95\textwidth]{plots/sift-knn-10.png}
\subcaption{Sift for $k=10$.    }
\label{fig:results:sift-scaling}
\end{subfigure}%
\begin{subfigure}[b]{0.47\textwidth}
\includegraphics[width=0.95\textwidth]{plots/radio-ml-knn-10.png}
\subcaption{RadioML for $k=10$ at SnR = 10dB.
}
\label{fig:results:radioml-scaling}
\end{subfigure}%
\vspace{1em}
\\
\begin{subfigure}[b]{0.47\textwidth}
\includegraphics[width=0.95\textwidth]{plots/silva-knn-10.png}
\subcaption{Silva for $k=10$.    }
\label{fig:results:silva-scaling}
\end{subfigure}%
\begin{subfigure}[b]{0.47\textwidth}
\includegraphics[width=0.95\textwidth]{plots/random-knn-10.png}
\subcaption{A random dataset for $k=10$.
}
\label{fig:results:random-scaling}
\end{subfigure}%

>>>>

Figures~\ref{fig:results:fashion-mnist-scaling}, ~\ref{fig:results:glove-25-scaling}, and ~\ref{fig:results:sift-scaling} show the scaling behavior of CAKES algorithms and existing algorithms on augmented versions of the following ann-benchmark datasets: fashion-mnist under Euclidean distance, glove-25 under cosine distance, and sift under Euclidean distance.
Figure~\ref{fig:results:random-scaling} shows the scaling behavior of CAKES algorithms and existing algorithms on a completely randomly generated dataset with same cardinality and dimensionality as sift (1 million points in 128 dimensions).
Figures~\ref{fig:results:silva-scaling} and ~\ref{fig:results:radioml-scaling} show the scaling behavior of CAKES on Silva and RadioML respectively.
The horizontal axis in each figure shows the cardinality of the augmented dataset with synthetic points (see Section \ref{subsec:methods:synthetic-data}).
The left-most point on each line is on the original dataset without any synthetic augmentation.
The vertical axis denotes throughput in queries per second.
Both axes are on a logarithmic scale.
In this section, we report results only for $k$-NN search with $k = 10$, but similar plots for $k=100$ can be found in the Supplement.
For HNSW and ANNOY, we report the recall for each measurement in the plots. For algorithms which exhibit recall greater than $0.9995$, we do not report the recall in the plots, but we do report it in the tables below.

Tables \ref{table:results:ann-fashion}, \ref{table:results:ann-glove-25}, \ref{table:results:ann-sift}, and \ref{table:results:ann-random} show the throughput and recall of CAKES's algorithms at each cardinality for  Fashion-Mnist, Glove-25, Sift, and a random dataset respectively. Though the scaling plots above present results for each of CAKES's three algorithms separately, the results in the CAKES column in these tables represent the fastest CAKES algorithm at that dataset and cardinality only. 
Given that we redo the auto-tuning process described in Section~\ref{subsec:methods:auto-tuning} for each augmented dataset, these values are an accurate representation of CAKES's performance on each dataset and cardinality multiplier. 
When reporting recall, we use 1.000* to denote that the recall is imperfect, but rounds to 1.000 when we consider only three decimal places.

In Figure~\ref{fig:results:fashion-mnist-scaling}, which shows results for Fashion-Mnist, we observe that as cardinality increases, the CAKES algorithms (GreedySieve in blue, Repeated $\rho$-NN in green, and Sieve in purple) are faster than na\"{i}ve linear search (plotted in orange).
In particular, we see that GreedySieve and Sieve begin to exhibit sublinear time performance at a cardinality near $10^5$ (i.e., at a multiplier of about {\color{red} add value} times the original dataset's cardinality). 
We also observe that CAKES's algorithms outperform FAISS-IVF (plotted in pink) for all cardinalities larger than the original dataset's cardinality.
Though state-of-the-art algorithms HNSW (plotted in gray), and ANNOY (plotted in yellow) are faster than CAKES for all cardinalities, we note that CAKES always exhibits perfect recall while HNSW and ANNOY exhibit much lower recall, as shown in Table~\ref{table:results:ann-fashion}.
For example, at a multiplier as low as eight, HNSW and ANNOY have recall of 0.525 and 0.857 respectively,
and this recall gets worse as the multiplier increases.
Meanwhile, CAKES exhibits perfect recall.
Amongst the CAKES algorithms, GreedySieve consistently performs best on this dataset, exhibiting sublinear time performance with increasing multipliers.

With glove-25 (Figure~\ref{fig:results:glove-25-scaling}), we observe that each of CAKES' algorithms outperforms linear search at some cardinality.
Specifically, all three of CAKES' algorithms outperform linear search at a cardinality just slightly more than the original dataset's cardinality, $10^6$.
Similar to the results for Fashion-Mnist, we see that CAKES's algorithms outperform FAISS-IVF for all cardinalities greater than $10^7$.
We see once again that HNSW and ANNOY are faster than CAKES's algorithms for all cardinality multipliers, CAKES's algorithms exhibit better recall at all multipliers, as shown in Table~\ref{table:results:ann-glove-25}.
In particular, at a multiplier as low as two, HNSW and ANNOY have recall as low as 0.607 and 0.832 respectively, and this recall continues to decrease as we increase the multiplier.
In contrast, CAKES exhibits perfect recall at all multipliers.
On this dataset, unlike with fashion-mnist, Repeated $\rho$-NN is consistently the best-performing CAKES algorithm.

In Figure~\ref{fig:results:sift-scaling}, which displays the scaling behavior of algorithms on the Sift dataset, we observe that while linear search performs best for lower cardinalities, our algorithms begin to outperform our implementation of linear search and FAISS-IVF at a cardinality near $10^7$.
For this dataset, Sieve, is the fastest CAKES algorithm for all cardinalities.
Similar to the results with Fashion-Mnist and Glove-25, we observe that while HNSW and ANNOY are faster than CAKES's algorithms for all cardinality multipliers, CAKES has better recall.
Table~\ref{table:results:ann-sift} shows that even with a multiplier of 1 (i.e., the original dataset), HNSW and ANNOY have recall as low as 0.782 and 0.686 respectively, while CAKES exhibits perfect recall.

With the random dataset, as seen in Figure~\ref{fig:results:random-scaling}, we observe that the three CAKES algorithms are the slowest out of all algorithms, at all cardinalities. 
As with the real datasets, HNSW and ANNOY are the fastest of any of the algorithms, and CAKES exhibits perfect recall at all cardinalities. 
HNSW and ANNOY exhibit \emph{much} lower recall on this random dataset than on the real datasets; 
in particular, with a multiplier of 1, HNSW and ANNOY have recall as low as 0.060 and 0.028 respectively, as reported in Table~\ref{table:results:ann-random}.

For Silva and RadioML, we benchmarked only CAKES's algorithms.
Due to the massive sizes of these datasets, we took random sub-samples of the dataset with lower cardinalities for our benchmarks, rather than augmented versions of the dataset.
With Silva, as shown in Figure~\ref{fig:results:silva-scaling}, we observe that for all algorithms, throughput seems to linearly decrease as cardinality increases, but that it seems to begin levelling off near cardinality $10^5$. Until cardinality near $10^4$, GreedySieve is the fastest CAKES algorithm, but for cardinalities greater than that, Repeated $\rho$-NN is the fastest CAKES algorithm. 
For RadioML, as shown in Figure~\ref{fig:results:radioml-scaling}, we observe that throughput declines nearly linearly the CAKES algorithms exhibit nearly indistinguishable performance for all cardinalities. In particular, throughput of CAKES algorithms is identical within three decimal places for all cardinalities.

\begin{figure}
\begin{subfigure}[b]{0.47\textwidth}
\includegraphics[width=0.95\textwidth]{plots/fashion-mnist-knn-10.png}
\subcaption{Fashion-Mnist for $k=10$.}
\label{fig:results:fashion-mnist-scaling}
\end{subfigure}%
\begin{subfigure}[b]{0.47\textwidth}
\includegraphics[width=0.95\textwidth]{plots/glove-25-knn-10.png}
\subcaption{Glove-25 for $k=10$.  }
\label{fig:results:glove-25-scaling}
\end{subfigure}%
\vspace{1em}
\\
\begin{subfigure}[b]{0.47\textwidth}
\includegraphics[width=0.95\textwidth]{plots/sift-knn-10.png}
\subcaption{Sift for $k=10$.    }
\label{fig:results:sift-scaling}
\end{subfigure}%
\begin{subfigure}[b]{0.47\textwidth}
\includegraphics[width=0.95\textwidth]{plots/radio-ml-knn-10.png}
\subcaption{RadioML for $k=10$ at SnR = 10dB.
}
\label{fig:results:radioml-scaling}
\end{subfigure}%
\vspace{1em}
\\
\begin{subfigure}[b]{0.47\textwidth}
\includegraphics[width=0.95\textwidth]{plots/silva-knn-10.png}
\subcaption{Silva for $k=10$.    }
\label{fig:results:silva-scaling}
\end{subfigure}%
\begin{subfigure}[b]{0.47\textwidth}
\includegraphics[width=0.95\textwidth]{plots/random-knn-10.png}
\subcaption{A random dataset for $k=10$.
}
\label{fig:results:random-scaling}
\end{subfigure}%
